???A quantitative framework for investigating the reliability of network construction

### 2017-09-18 manuscript/CI_interactions.tex
fixed a merge

@@

Starting from the problem described in the previous section, we consider that a pair of species have been observed $n$ times co-occurring at one location, and the number of times they interact is $k = 0$. We aim to evaluate the uncertainty of this interaction. What this is mean exactly ? We consider there are three nested level of uncertainty making this phenomenon a stochastic process :

\section*{Maximum likelihood estimate and confidence interval of interaction probability}

Estimation of the detection and process uncertainties could be tedious, if not infeasible in some instances. As a consequence, for now, we will focus on the estimate of the interaction probability $\lambda$, assuming that both the detection and process probabilities, $\delta$ and $\chi$ respectively, are equal to 1. The occurrence of an interaction is a Bernoulli trial. Consequently, the number of success $k$ over $n$ trials will follow a binomial distribution:

@@

The situation where $k = 0$ is also problematic because we have very few information to estimate $\lambda$, in particular when $n$ is also small. We therefore propose a Bayesian approach to solve this problem, building on two relevant features of such statistics : posterior distribution of parameter estimates and usage of prior information.

@@

\section*{A quantitative example}
%Perhaps we want to also have a trait matching example in addition to the degree distribution example below?
The framework can be illustrated with a simple quantiative example. Suppose we have $n = 10$ observations of co-occurrence between speices $i$ and species $j$ in a given time interval and area, and $X = 2$ observations of interactions. The maximum likelihood estimate of the interaction probability is simply $\theta_{MLE} = 2/10 = 0.2$. 
% We should probabaly have a value  on the interaction probablitity that differs from the average of the degree distrbution
 Now consider we know that species $i$ is known to interact with 10 other species, which have the following standardized degrees:

\vspace{12pt}
\noindent\emph{
    degree = c(0.66, 0.09, 0.06, 0.08, 0.04, 0.30, 0.06, 0.38, 0.03, 0.19)
    }.
  \vspace{12pt}

Using these standardized degrees to inform our prior expectation of the 
distribution of interaction probabilities and the R code from the 
connectance example, we obtain prior parameters $\alpha$=0.9034876 
and $\beta$=3.6623995. Using these priors in equations~\ref{mean} 
and~\ref{variance} above, we find an average interaction probability of 0.1993348 and a variance of 0.01025322 [[Different from Dom's values]].
% n=number of observed co-occurances.
The average interaction probability is 0.2163971 and the variance is 0.02353319. The hyper parameters are computed according to the following script, using the function \emph{fitdistr} from the R~\citep{R} package MASS~\citep{MASS}, yielding:

\vspace{12pt}
\noindent\emph{    pars = fitdistr(x = degree, "beta", start = list(shape1 = 1, shape2 = 1),lower=c(0,0))\$estimate }.
  \vspace{12pt}

The lower bound of (0,0) for the two parameters of the Beta distribution prevents the function from attempting to fit invalid (negative) parameter estimations.

The average of the distribution is computed as follows:

  \vspace{12pt}
\noindent
\emph{
    \noindent n = 10
    \\\noindent k = 2
    \\\noindent alpha\_prime = alpha + k
    \\\noindent beta\_prime = beta + n - k
    \\\noindent mean\_theta = alpha\_prime/(alpha\_prime+beta\_prime)
},
  \vspace{12pt}

which yields a value $E[\theta]=0.2057651$ [[I get 0.1993348]]. And the variance:

  \vspace{12pt}
\noindent\emph{alpha\_prime$\times$ beta\_prime\/(alpha\_prime+beta\_prime)$^2$\/(alpha\_prime+beta\_prime+1)},
  \vspace{12pt}

which yiealds a value $var[\theta] = 0.009579623$. [[I get 0.01025322]]

The credible interval is not symmetric around the mean since the distribution is bounded between 0 and 1 and is not symmetric. In this case, a 95\% credible interval for the mean of theta is (0.0448479, 0.4308686) and can be computed using the quantile function \emph{qbeta}:

  \vspace{12pt}
\noindent\emph{
    \noindent lowCI = qbeta(p = 0.025, shape1 = alpha\_prime, shape2 = beta\_prime)
    \\\noindent highCI = qbeta(p = 0.975, shape1 = alpha\_prime, shape2 = beta\_prime)}.
  \vspace{12pt}

Now, consider the case where the two species have never been observed interacting across $n$ trials. The question is then ``what is the probability these two species do not interact''? Since it is not possible to prove that the two species could never interact, given the framework above we must fix a threshold below which we consider that there is no interaction. We call this threshold probability $\theta*$. We then use the cumulative distribution function to estimate $P(\theta<\theta*|X=0,n)$. The following script illustrates this probability for an increasing number of trials and $\theta*$=0.1:

  \vspace{12pt}
\noindent\emph{
    \noindent n = seq(0,100,1)
    \\\noindent X = 0
    \\\noindent theta\_star = 0.1
    \\\noindent cdf = pbeta(theta\_star, shape1 = alpha+X, shape2 = beta+n-X)
    \\\noindent plot(n,cdf, type = "l", xlab = "n", ylab = "Cumulative distribution")
    \\\noindent abline(h = 0.95,lty = 3)}.
  \vspace{12pt}

The exercise exemplified above yields a surprising result: it requires \textgreater30 [[or 23??]] observations of no interactions to be 95\% sure that the interaction probability is smaller than 0.1. Note the special case where there is no observation of the two species co-occurring and failing to interact, $n = 0$. In this situation, the posterior distribution converges to the prior distribution.

@@

      [[add references to formulae]]

      For species where no co-occurances were observed, we can calculate the MLE estimates for the mean and variance of $\lambda_{ij}$ directly from this prior. These are:
      $\hat{\lambda_{ij}}$=0.5451437, var($\lambda$)=0.0481417.
      This is consistent with the prior data, where the average interaction probability based on the degree distributions of Salix genotypes and gallers was 0.541. [[What's the connectance of our dataset, of we call greys 0?]] This probability does, however, seem rather high for our dataset. This discrepancy may be because~\citep{Balbour2016} were considering genotypes of a single \emph{Salix} species and only four gallers from a single family rather than multiple \emph{Salix} species and galler families as is the case in our dataset.

>>>>

\section*{A naive qunatification of interaction probability and uncertainty}

@@

As a start, we will consider how we could naively quantify intetraction probability and its associated uncertainty. We consider that a pair of species have been observed $n$ times co-occurring at one location, and the number of times they interact is $k = 0$. We aim to evaluate the uncertainty of this interaction. The occurrence of an interaction is a Bernoulli trial. Consequently, the number of success $k$ over $n$ trials will follow a binomial distribution:

@@

The situation where $k = 0$ is also problematic because we have very few information to estimate $\lambda$, in particular when $n$ is also small. We therefore propose a Bayesian approach to solve this problem, building on two relevant features of such statistics : posterior distribution of parameter estimates and usage of prior information. But before moving to this direction, we will first take time to better describe what is exactly meant by an interaction probability and appropriately define sources of uncertainty.

So many sources of interaction pooled into a single entity; hard to estimate.

MLE is zero if no interactions detected, but there may still be a substantial probability of interaction should the two species co-occur. [This is particularly important in the context of a changing world, where species not co-occurring today – or species not detected co-occurring today – may well do so tomorrow, forming complex interactions.]

What this is mean exactly ? We consider there are three nested level of uncertainty making this phenomenon a stochastic process :

For the ecologist working on collecting network data, and on later interpreting these data, the considerations above raise a major challenge: how may we infer whether unobserved interactions went undetected due to sampling, or whether they truly do not occur. How then may we refine our sampling approaches to reduce uncertainties, and to gain insights into the impact of multiple processes on field observations? Importantly, some sources can be minimized with appropriate sampling design and efforts, while other sources are difficult or impossible to reduce – since they are generated by chance variation created by the very process in which we are interested.

The basic information that we are collecting consists of interaction matrices of interaction partners. If our counts of whether species A is interacting with species B reflects both uncertainty with respect to whether two species interact when co-occurring at a given site, chance variation in whether the interaction is realized, and uncertainty with respect to whether it is detected when realized, then what can we do to pin down any of these different parts?

A simple consideration will illustrate the problem: assuming that we want to infer the probability of an interaction between two species, A and B. Now assume that in reality, interaction between A and B is plain impossible (i.e. the true (P(L|co-occurrence of A and B = 0). Given no added detection error in observing the incidence of the interaction, then a single observation of species co-occurrence will reveal (very) little regarding the probability of the interaction – the confidence limits for one observation essentially spans from 0 to 1 (see following section). For 5 observations, this confidence limit will still extends to 0.5. Only with thirty observations will it shrink to 0.1. Thus, adding (many) more observations is certainly useful in controlling uncertainty due to stochasticity inherent in the process itself.

The pattern of Fig. X (binomial CL fig from Dom) should be compared to the frequency of co-occurrence of taxa in the Salix-galler data set. Here, most(TO COMPUTE) pairs of species are never found co-occurring, and for species found co-occurring the number of co-occurrences was generally low (Fig from AC and KW). Among the species which did sometimes interact when co-occurring, the incidence of interaction was actually low(TO COMPUTE) (freq histogram Fig from AC and KW). Thus, even in the most extensive data set that we could think of, limited sampling is clearly a major source of uncertainty.

The obvious rule of thumb derived from the above example will be “sample more”, but we note that there are limits to the utility of this motto. Sampling more will clearly reduce uncertainty regarding the upper bound of the probability of interaction, and it will also increase the probability of detecting unlikely interactions (i.e. interactions occurring with a low probability). Yet, since the probability of observing the co-occurrence of two species will always be higher than the probability of observing their interaction (since the probability of interaction is conditional on both interaction partners being present), we will accumulate observations of co-occurrences quicker than we will accumulate observations of interactions. Thus, the more we sample, the more holes will appear in our interaction matrix, which may seem like a counter-intuitive outcome.

As a solution towards deciding whether unobserved interactions were undetected due to sampling, or whether they truly do not occur, Weinstein et al. (2017) used daily repeated sampling rounds to estimate the daily probability of detecting a hummingbird interaction, and to thereby model the observation and process mechanisms (Weinstein, B. G., & Graham, C. H. (2017). On comparing traits and abundance for predicting species interactions with imperfect detection. Food Webs 11; 17-25: https://doi.org/10.1016/j.fooweb.2017.05.002). While conceptually attractive, this approach is unsuitable for interactions occurring over longer time scales (like associations between hosts and parasitoids with a single generation per year), or very rare interactions (since no matter how frequently and extensively we sample, there will still be some rare interactions lurking around). What is worse, the problem persists that if a given interaction is not observed on a given day, then it may still be either because it was impossible that day ((P(L|co-occurrence of A and B)=0), because interaction between the two did occur but was not observed, or from any combination of the two. Thus, from a conceptual perspective, this approach will fail to satisfactorily separating between all sources of uncertainty. An interaction not observed on day t may have gone unobserved either because there were constraints on the process (i.e. the probability was really zero), because of chance variation in the realized incidence of an event associated with a finite probability. Most importantly, if during a series of days, two species are simply never observed co-occurring, then we still have not learnt a thing about their probability of interacting should they do so. In other words, if the two interaction partners do not co-occur, we will still gain no information.

An added complication is that not all sourced of uncertainty are proportional to sample size. To score an interaction between A and B, we will need to identify both partners correctly (a non-trivial problem in many food webs; e.g. Kaartinen et al. 2011 Ecol Entomol; Roslin and Majaneva 2016) and be able to resolve all interactions with a similar likelihood. For both molecular (ref ref) and rearing techniques, given types of interactions may go unnoticed due to technical challenges.

The bottom line is that separating between different sources of uncertainty is difficult indeed. In the best of all possible worlds, we might then use an experimental approach to estimate detection uncertainty: we might create replicate sets of cases where the interaction is really there, then evaluating them and examining the fraction of cases scoring positive. Yet, as anyone attempting these experiments will know, most interaction partners are proving notoriously unwilling to collaborate. Gallers refuse to lay eggs on their host plants (ref ref), parasitoids to attack their hosts when experimentally introduced (Gariepy 2017; ref ref). If we introduce only one of the components (like in the cases where we expose a set of hosts to parasitism; ref ref; Gripenberg et al. xxx) then we are already confounding multiple levels of uncertainty in the outcome – that of interaction uncertainty (whether the species will interact once both there), process uncertainty (whether the interaction will occur if and since the probability of interaction is below one), and detection uncertainty (i.e. whether we will observed the interaction once realized). To perform these experiments in a comprehensive and reliable manner, we will then need to expose a large number of interactions in a comprehensive experiment across space. This type of “cafeteria experiments in space” is not only unfeasible, but also undermining our original objective: If the local occurrence of interactions is uninformative regarding the underlying probabilities of occurrence, then we should be doing something else than wasting a monumental effort on collecting local interaction matrices. <deep sigh and forget that section? >

As a tentative solution to disentangling observation and process mechanisms, we propose that some insights regarding the detectability of interactions between species not found co-occurring may be gained from data on other species, for which it is known that the interaction is feasible. If we use this information obtain the joint probability of detection and process (e.g. to glean information regarding at how many sites the species are found interacting across the entire dataset), and assuming that most of this pattern is due to variation in detection rate, then insights into the incidence of interactions between species may be gleaned from the accumulation rate of positive interactions at sites were interaction is possible (i.e. with at least one observation of a positive interaction) [this reasoning to be further formalized]

@@

\section*{A quantitative example}
%Perhaps we want to also have a trait matching example in addition to the degree distribution example below?
% Are degree distributions valid priors for interaction probabilities in a unipartite web? For the Salix-gall dataset we're using interaction probabilities as priors (since the degree of both Salix and galler matter)... maybe we should rephrase this to predict the degree of species $i$ based on the degrees of 10 other species in the network?
The framework can be illustrated with a simple quantiative example. Suppose we have $n = 10$ observations of co-occurrence between species $i$ and species $j$ in a given time interval and area, and $X = 3$ observations of interactions. The maximum likelihood estimate of the interaction probability is simply $\theta_{MLE} = 3/10 = 0.3$. 
% We should probabaly have a value  on the interaction probablitity that differs from the average of the degree distrbution
 Now consider we know that species $i$ is known to interact with 10 other species, which have the following standardized degrees:

\vspace{12pt}
\noindent\emph{
   degree = c(0.66, 0.09, 0.06, 0.08, 0.04, 0.16, 0.06, 0.38, 0.03, 0.02)
    }.
  \vspace{12pt}

The mean of these standardized degrees is 0.158, approximately half the $\theta_{MLE}$ obtained from the observed data. We can use these standardized degrees to inform our prior expectation of the 
distribution of interaction probabilities. With some simple R code (function calculate_parameters, \emph{Appendix S2}), we obtain prior parameters $\alpha$=0.7559544
and $\beta$=3.6710277. Using these priors in equations~\ref{mean} 
and~\ref{variance} above (or in the R function calculate_distribution in \emph{Appendix S2}), we find a prior mean interaction probability of 0.1707607 and a prior variance of 0.0319860.

Adding the observed data (n=10, k=3) and using the same code,
we obtain posterior hyper parameters $\alpha'$=3.755954 and $\beta'$=10.671028 and a posterior mean 0.2603423 and variance 0.0133475 for the probability of. Comparing the posterior distribution to the prior, we see that the posterior is closer to the observed data and that the additional data has reduced the variance.

We may also wish to calculate a credible interval (analogous to the frequentist confidence interval). This is also quite straightforward (see function credible_interval in \emph{Appendix S2}). In this case, a 95\% credible interval for the mean of theta is (0.07755119, 0.50557509).

Now, consider the case where the two species have never been observed interacting across $n$ trials. The question is then ``what is the probability these two species do not interact''? Since it is not possible to prove that the two species could never interact, given the framework above we must fix a threshold below which we consider that there is no interaction. We call this threshold probability $\theta*$. We then use the cumulative distribution function to estimate $P(\theta<\theta*|X=0,n)$. The 
function plot_precision in \emph{Appendix S2} calculates distribution function for $\theta*$ with an increasing number of trials. The figure below illustrates this distribution for $\theta*$=0.1.

The exercise exemplified above yields a surprising result: it requires \textgreater21 observations of no interactions to be 95\% sure that the interaction probability is smaller than 0.1. Note the special case where there is no observation of the two species co-occurring and failing to interact, $n = 0$. In this situation, the posterior distribution converges to the prior distribution.

@@

      For species where no co-occurances were observed, we can 
      calculate the MLE estimates for the mean and variance of 
      $\lambda_{ij}$ directly from this prior following 
      equations~\ref{mean} and~\ref{variance} 
      (see \emph{Appendix S2} for R implementation). The prior 
      distribution we obtain based on~\citet{Balbour2016} is:
      $\hat{\lambda_{ij}}$=0.545, var($\lambda$)=0.048.
      This is consistent with the prior data, where the average 
      interaction probability based on the degree distributions of Salix genotypes and gallers was 0.541. [[What's the connectance of our dataset, if we call greys 0?]] This probability does, however, seem rather high for our dataset. This discrepancy may reflect the differences in scale between a network containing different genotypes of a single \emph{Salix} species and four gallers from a single family and a network containing many \emph{Salix} species and gallers from many families as is the case in our dataset.
